---
title: "Supplemental Material 5: Interaction Data Analysis Report"
# author: "Amish Mishra"
date: "December 19, 2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(DescTools)

# # Set global theme with Arial font and 10pt font size - not necessary
# theme_set(
#   theme_gray(base_family = "Arial", base_size = 10) +
#     theme(text = element_text(family = "Arial", size = 10))
# )
```

This file reproduces the figures and statistics reported in the manuscript for quantifying the interaction effects observed. As a guide, this table helps map variable names in the code to the biological meanings:

|Code Category Name| Code Variable Name | Biological Variable Name |
|------------------|--------------------|--------------------------|
|   `new_gene`     | `zip/+`            | $zip^1/+$                |
|                  | `sqh/+`            | $sqh^{AX3}/+$            |
|------------------|--------------------|--------------------------|
|                  | `+/+`              | $Gli^+/Gli^+$                   |
|                  | `dv5/+`            | $Gli^{dv5}/+$            |
|   `dv5_base`     | `dv5/RAR`          | $Gli^{dv5}/Gli^{RAR77}$  |
|                  | `dv5/dv5`          | $Gli^{dv5}/Gli^{dv5}$    |


# Load necessary libraries and do setup
```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(FSA)
library(latex2exp)

# Set bootstrap samples
NUM_BOOTSTRAPS = 10000
```

# Get data ready

## Load and clean data
```{r, warning=FALSE}
raw_data <- read_excel("data/interactionscores.xlsx", 
    col_types = c("skip", "numeric", "numeric",
                  "skip", "numeric", "numeric", "numeric",
                  "skip", "numeric", "numeric", "numeric",
                  "skip", "numeric", "numeric", "numeric"))

# Select the first 90 rows
WingScores_by_cols <- head(raw_data, 90)

str(WingScores_by_cols)
head(WingScores_by_cols)
```

## Pivot data into long format

```{r}
WingScores_long <- WingScores_by_cols %>%
  pivot_longer(
    cols = everything(),         # Specify all columns to be pivoted
    names_to = "mutation",       # Name of the new column for variable names
    values_to = "score"          # Name of the new column for scores
  ) %>%
  drop_na()

dv5_levels <- c("+/+", "dv5/+", "dv5/RAR", "dv5/dv5")
WingScores <- WingScores_long %>%
  separate(mutation, into = c("dv5_base", "new_gene"), sep = "&", convert = TRUE) %>%
  mutate(dv5_base = trimws(dv5_base),
         new_gene = trimws(new_gene),
         dv5_base = factor(dv5_base, levels = dv5_levels, ordered = TRUE),
         new_gene = as.factor(new_gene))

str(WingScores)
head(WingScores)
summary(WingScores)
```

\newpage

# Explore
This section gives some summary statistics and initial visualizations of the dataset.

## Counts of scores (how many samples for each crossed category)
```{r}
count_table <- WingScores %>%
  count(dv5_base, new_gene)

# Convert the count table into a two-way table
counts_two_way_table <- count_table %>%
  pivot_wider(names_from = new_gene, values_from = n)

rownames(counts_two_way_table) <- NULL
counts_two_way_table <- counts_two_way_table %>% column_to_rownames(var = "dv5_base")

print(counts_two_way_table)
```

## Medians of score
```{r}
# Summarize the data to get the median score for each combination of dv5_base and new_gene
median_scores <- WingScores %>%
  group_by(dv5_base, new_gene) %>%
  summarize(median_score = median(score, na.rm = FALSE))

med_summary_table <- xtabs(median_score ~ dv5_base + new_gene, data = median_scores)
med_summary_table['+/+', '+/+'] <- NA
print(med_summary_table)
```

## Means of score
```{r}
# Summarize the data to get the mean score for each combination of dv5_base and new_gene
mean_scores <- WingScores %>%
  group_by(dv5_base, new_gene) %>%
  summarize(mean_score = mean(score, na.rm = TRUE))

mean_summary_table <- xtabs(mean_score ~ dv5_base + new_gene, data = mean_scores)
mean_summary_table['+/+', '+/+'] <- NA
print(mean_summary_table)
```

## Plots of score distributions

The histograms show the skewness of the data and indicate issues of normality of the distribution of scores. Since the data is measured on an ordinal scale with 7 possible values and does not satisfy the conditions of normality necessary to conduct ANOVA, we opt for focusing on the medians and using non-parametric test alternatives for assessing significance.

The dashed lines in the histograms are the medians for the respective distribution.

```{r fig.height=6, fig.width=8}
median_scores <- WingScores %>%
  group_by(dv5_base, new_gene) %>%
  summarize(median_score = median(score, na.rm = TRUE))

# Histogram with two factors
ggplot(WingScores, aes(x = score, fill = dv5_base)) +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "dodge") +
  facet_wrap(dv5_base ~ new_gene, nrow = 4, ncol = 3) +
  # theme_minimal() +
  labs(title = "Histogram of Scores by new_gene, Faceted by dv5_base",
       x = "Score",
       y = "Count") +
  scale_fill_discrete(name = "dv5_base") +
  geom_vline(data = median_scores, aes(xintercept = median_score), 
             color = "blue", linetype = "dashed", linewidth = 1)
```

These boxplots clearly show differences in medians between levels within the same category. They also show a spread of scores in each category. The means on these plots are marked with an "X".

```{r}
# Combine group and level into a single variable for x-axis
WingScores$x_combined <- paste(WingScores$dv5_base,
                                        WingScores$new_gene, sep = ";")

# Make labels for the boxplot x-axis
custom_labels <- c(
  "+/+;sqh/+"=TeX("$\\textit{sqh}^{AX3}/+$"),
  "+/+;zip/+"=TeX("$\\textit{zip}^1/+$"),
  "dv5/+;+/+"=TeX("$\\textit{Gli}^{dv5}/+$"), 
  "dv5/+;sqh/+"=TeX("$\\textit{sqh}^{AX3}/+$ ; $\\textit{Gli}^{dv5}/+$"),
  "dv5/+;zip/+"=TeX("$ \\textit{zip}^1$ , $\\textit{Gli}^{dv5}/+$"),
  "dv5/dv5;+/+"=TeX("$\\textit{Gli}^{dv5}/\\textit{Gli}^{dv5}$"), 
  "dv5/dv5;sqh/+"=TeX("$\\textit{sqh}^{AX3}/+$ ; $\\textit{Gli}^{dv5}/\\textit{Gli}^{dv5}$"),
  "dv5/dv5;zip/+"=TeX("$\\textit{zip}^1$ , $ \\textit{Gli}^{dv5}/\\textit{Gli}^{dv5}$"),
  "dv5/RAR;+/+"=TeX("$\\textit{Gli}^{dv5}/\\textit{Gli}^{RAR77}$"), 
  "dv5/RAR;sqh/+"=TeX("$\\textit{sqh}^{AX3}/+$ ; $\\textit{Gli}^{dv5}/\\textit{Gli}^{RAR77}$"),
  "dv5/RAR;zip/+"=TeX("$\\textit{zip}^1$ , $\\textit{Gli}^{dv5}/\\textit{Gli}^{RAR77}$")
  )

# Boxplot
p <- ggplot(WingScores, aes(x = x_combined, y = score, fill = new_gene)) +
  geom_boxplot(position = position_dodge(width = 1.0), width = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "black", 
                 position = position_dodge(width = 1.0)) +
  facet_wrap(~ dv5_base, scales = "free_x", nrow=1) +
  scale_fill_manual(values = c("white","lightgrey","darkgrey")) +
  scale_x_discrete(labels = custom_labels) +
  labs(title = "Side-by-Side Box Plots", x = NULL,
       y = "Hair alignment score", fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  guides(fill=FALSE) + 
  scale_y_continuous(
  breaks = seq(0, 7, by = 1), # Setting breaks for every unit
  labels = seq(0, 7, by = 1),  # Corresponding labels for the breaks
  minor_breaks = NULL
  )
# Save the plot
# ggsave("figures/hair_scores_boxplots_scaled.tiff", plot = p, width = 10, height = 5, dpi = 300)
p
```

# Analysis
Through investigation, we found that trying to use two-way ANOVA with interaction leads to failure of the assumptions of normality of residuals and homogeneity of variance of residuals. Thus, we have to resort to non-parametric methods.

## Kruskal-Wallis Test

A non-parametric test to see if there is a significant difference in the distributions of scores across the 3 levels of the `new_gene` factor for each level of the `dv5_base` factor. The following code and output shows that for each of the 4 levels of Gli, there is a significant difference in the distributions of the scores between the 3 `new_gene` levels.

```{r}
# Filter data for $Gli^+/Gli^+$ 
gli0_level_data <- WingScores %>% filter(dv5_base == '+/+')
kruskal_test <- kruskal.test(score ~ new_gene, data = gli0_level_data)
print(kruskal_test)

# Filter data for dv5/+
gli1_level_data <- WingScores %>% filter(dv5_base == 'dv5/+')
kruskal_test <- kruskal.test(score ~ new_gene, data = gli1_level_data)
print(kruskal_test)

# Filter data for dv5/RAR
gli2_level_data <- WingScores %>% filter(dv5_base == 'dv5/RAR')
kruskal_test <- kruskal.test(score ~ new_gene, data = gli2_level_data)
print(kruskal_test)

# Filter data for dv5/dv5
gli3_level_data <- WingScores %>% filter(dv5_base == 'dv5/dv5')
kruskal_test <- kruskal.test(score ~ new_gene, data = gli3_level_data)
print(kruskal_test)
```
## Dunn's test for pairwise comparisons (post-hoc analyses)

The Dunn's test adjusted p-values are the ones we report in the manuscript (annotated on the boxplot). Below, we check within each `dv5_base` category which pairs show significant differences in their medians using Dunn's test. The values we use are the ones in the column labeled `P.adj`.

### When `dv5_base = +/+`
```{r}
dunnTest(score ~ new_gene, data=gli0_level_data, method="bonferroni")
```
For $Gli^+/Gli^+$, there is no significant difference in the medians of the scores when the `new_gene` is `+/+` or when it is `zip/+`. However, the other two pairs show a significant difference in median scores.

### When `dv5_base = dv5/+`

```{r}
dunnTest(score ~ new_gene, data=gli1_level_data, method="bonferroni")
```
For the `dv5/+` Gli level, there is no significant difference in the medians of the scores when the `new_gene` is `sqh/+` or when it is `zip/+`. However, the other two pairs show a significant difference in median scores.

### When `dv5_base = dv5/RAR`
```{r}
dunnTest(score ~ new_gene, data=gli2_level_data, method="bonferroni")
```
For the `dv5/RAR` Gli level, there is no significant difference in the medians of the scores when the `new_gene` is `sqh/+` or when it is `zip/+`. However, the other two pairs show a significant difference in median scores.

### When `dv5_base = dv5/dv5`

```{r}
dunnTest(score ~ new_gene, data=gli3_level_data, method="bonferroni")
```
For the `dv5/dv5` Gli level, there is no significant difference in the medians of the scores when the `new_gene` is `sqh/+` or when it is `zip/+`. However, the other two pairs show a significant difference in median scores.

## Bootstrapping for finding the signficance of the interactions

The strategy is described as follows:

1. Generate a stratified bootstrap sample for each of the `dv5_base` levels with no `new_gene` mutation. 
2. Generate a stratified bootstrap sample for each of the `new_gene` mutations (`sqh/+` and `zip/+`) where `dv5_base = +/+`.
3. Add the response values for each cross of the `new_gene` and `dv5_base` levels to get the expected additive value. Only compute as many added values as there were in the original dataset. So, for example, the original dataset had 18 values for `dv5/dv5` with `sqh/+` so only 18 rows will be computed in the bootstrap sample.
4. Find the median of each crossed-level.
5. Repeat this procedure 10,000 times to generate a null distribution.
6. Calculate a p-value as the number of bootstrap samples with median values greater than or equal to the observed value of the crossed levels of the original data. There will be 6 p-values to compute since there are 3 levels of interest of the `dv5_base` and 2 levels of interest of the `new_gene`. 

<!-- The format for interaction is $(a,x)$ vs. $(b,y)$ where $a$ and $b$ are levels of the `dv5_base` factor and $x$ and $y$ are levels of the `new_gene` factor. If there was no interaction, then the change from $x$ to $y$ for $a$ should be the same as the change from $x$ to $y$ for $b$. -->

<!-- In the code, the variable `diff_gli2_sqh` (for example) finds the change in the medians between the median score for `dv5/RAR` and the median score for `dv5/RAR & sqh/+`. We then compute the difference in medians of $Gli^+/Gli^+$ with `sqh/+` data and $Gli^+/Gli^+$ with no mutation data and store that as the variable `diff_gli0_sqh`. Finally, the interaction is computed as the difference `diff_gli2_sqh - diff_gli0_sqh`. If this value is 0, then there is no interaction between the two factors at these levels. If this value is non-zero, then there is evidence for interaction. -->

The analysis in this section helps to establish how significant is the more-than-additive effect we see of combining, for example, `sqh/+` with `dv5/RAR` (resulting in a score increase of 2 when 0 is expected).

### Function to calculate the expected medians of combining base levels from each factor
```{r}
# Define the bootstrapping function
bootstrap_expected <- function(data, counts_two_way_table, n_bootstrap = NUM_BOOTSTRAPS) {
  set.seed(123) # For reproducibility
  boot_expected <- replicate(n_bootstrap, {
    
    # Resample the data stratified
    resample_data <- data %>%
      group_by(dv5_base, new_gene) %>%
      slice_sample(prop = 1, replace = TRUE) %>%
      filter((dv5_base == "+/+" & new_gene != "+/+") | 
              (dv5_base != "+/+" & new_gene == "+/+")) %>%
      ungroup()
    
    # Calculate the expected additive effects
    # Loop over the vectors and filter data
    gli_vals <- c("dv5/+", "dv5/RAR", "dv5/dv5")
    new_gene_vals <- c("sqh/+", "zip/+")
    
    expected_data <- data.frame()
    for (i in gli_vals) {
      for (j in new_gene_vals) {
        # Filter the data by corresponding values in gli_vals and new_gene_vals
        gli_only <- resample_data %>%
          filter(dv5_base == i, new_gene == '+/+')
      
        new_gene_only <- resample_data %>%
          filter(dv5_base == '+/+', new_gene == j)
        
        # Only add as many values together as there were in the original dataset for the crossed levels
        num_to_compute <- counts_two_way_table[i, j]
        sum <- head(gli_only, num_to_compute)$score + head(new_gene_only, num_to_compute)$score
        expected <- data.frame(dv5_base = rep(NA, num_to_compute),
                              new_gene = rep(NA, num_to_compute),
                              score = rep(NA, num_to_compute))
        expected$score <- sum
        expected$dv5_base <- i
        expected$new_gene <- j
        expected_data <- bind_rows(expected_data, expected)
      }
    }
    resample_data <- bind_rows(resample_data, expected_data)
    
    # Compute average for the resampled data
    resample_medians <- resample_data %>%
      group_by(dv5_base, new_gene) %>%
      summarize(average_response = median(score), .groups = 'drop') %>%
      filter((dv5_base != "+/+" & new_gene != "+/+"))

    # Combine expected into a vector
    boot_expected <- resample_medians$average_response
  })
  return(boot_expected)
}

# Perform the bootstrapping
boot_results <- bootstrap_expected(WingScores, counts_two_way_table, n_bootstrap = NUM_BOOTSTRAPS)

# Convert bootstrapped differences to a long format data frame for ggplot
boot_expected_long <- as.data.frame(t(boot_results)) %>%
  pivot_longer(cols = everything(), names_to = "Comparison", values_to = "Expected")

# Create labels for the comparisons
comparison_labels <- c("gli1_sqh_expected", "gli1_zip_expected", 
                        "gli2_sqh_expected", "gli2_zip_expected", 
                        "gli3_sqh_expected", "gli3_zip_expected"
  )

# Add labels to the data frame
boot_expected_long <- boot_expected_long %>%
  mutate(Comparison = factor(Comparison, labels = comparison_labels))
```

### Visualize
The plot below shows the distribution of the expected medians of the crossed levels of various bootstrap samples. The dashed red lines in each subplot indicate the observed median score for the respective combination of `dv5_base` and `new_gene` from the original data.

<!-- 
The language may seem tough to parse, but the story follows the same theme as shown in the boxplots from earlier. We saw that the difference in median scores between, for example, `sqh/+` and `+/+` with $Gli^+/Gli^+$ was 0 but when considering the difference in median scores between the `dv5/RAR` with `sqh/+` and the `dv5/RAR` without a mutation, the difference in medians was 2. This was visually evident by how much farther up the lightgray boxplot in the `dv5/RAR` category was as compared to the white boxplot in the same category. -->

<!-- # Calculate the median of bootstrapped interactions for each comparison -->
<!-- ```{r} -->
<!-- boot_medians <- boot_expected_long %>% -->
<!--   group_by(Comparison) %>% -->
<!--   summarize(median_difference = median(Expected)) -->
<!-- ``` -->


```{r fig.height=3.5, fig.width=7}
# Save the true averages in a dataframe
true_vals <- data.frame(Comparison = comparison_labels, 
                        true_scores = as.vector(t(med_summary_table[2:4, 2:3])))

# Histogram of bootstrapped differences
ggplot(boot_expected_long, aes(x = Expected, fill = Comparison)) +
  geom_histogram(binwidth = 0.5, alpha = 0.7, position = 'identity') +
  geom_vline(data = true_vals, aes(xintercept = true_scores),
             color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ Comparison, scales = "free",  nrow = 4, ncol = 2) +
  theme_minimal() +
  labs(title = "Distribution of Bootstrapped Medians",
       x = "Expected Medians",
       y = "Frequency")
```
To compute p-values for the interaction, we compute the proportion of interaction values that were greater than or equal to the true (observed) median score value for the respective combination of `dv5_base` and `new_gene`. The p-values can be interpreted in the standard sense: the probability of obtaining this evidence by chance against the null hypothesis. The null hypothesis is that the effect of the `new_gene` (mutation) and `dv5_base` (Gli) is additive. Thus, a low p-value suggests that our evidence is strong against the null hypothesis and the effects we see in the respective combination of Gli and mutation is more than additive.

```{r}
merged_expected_true <- merge(true_vals, boot_expected_long, by = 'Comparison')

results <- merged_expected_true %>%
  group_by(Comparison) %>%
  summarise(num_extreme = sum(Expected >= true_scores),
            prop_extreme = mean(Expected >= true_scores))

print(results)
```

To read the table above, consider the row with "Comparison"=`gli3_zip_expected`. The `num_extreme =``r results[results$Comparison == "gli3_zip_expected", ]$num_extreme`, which tells how many of the bootstrap samples had a median score equal to or greater than the true value for that group from the original data.

The p-values for the significance of this result is given in the column titled `prop_extreme`.
