---
title: "Supplemental Material 4: RNAi Data Analysis Report"
# author: "Amish Mishra"
date: "May 30, 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(DescTools)

# # Set global theme with Arial font and 10pt font size - not necessary
# theme_set(
#   theme_gray(base_family = "Arial", base_size = 10) +
#     theme(text = element_text(family = "Arial", size = 10))
# )
```

This file performs the statistical analyses comparing the percentage of wing hair misalignments (termed "score") for varying temperatures of various genotypes.

# Data description

The dataset columns have the following descriptions.

| Column Name        | Description                                      |
|--------------------|--------------------------------------------------|
| `sample_ID`         | Unique identifier for each sample                   |
| `genotype`           | Genotype classification                          |
| `temp`               | Temperature (18, 22, 25, 29) measured in degrees celsius              |
| `score`           | Percentage of misaligned hairs in a 50 hair region                       |

# Genotype description

| Genotype             | Description                         |
|----------------------|------------------------------------|
| `bx-Gal4`            |        bx-Gal4                |
| `lacZ`               |    UAS-lacZ     |
| `sqh-RNAi`           |    UAS-sqh-RNAi   |
| `zip-RNAi`           |  UAS-zip-RNAi |
| `nub-Gal4`           |  nub-Gal4   |





<!-- |Code Category Name| Code Variable Name | Biological Variable Name | -->
<!-- |------------------|--------------------|--------------------------| -->
<!-- |   `new_gene`     | `zip/+`            | $zip^1/+$                | -->
<!-- |                  | `sqh/+`            | $sqh^{AX3}/+$            | -->
<!-- |------------------|--------------------|--------------------------| -->
<!-- |                  | `+/+`              | $Gli^+/Gli^+$                   | -->
<!-- |                  | `dv5/+`            | $Gli^{dv5}/+$            | -->
<!-- |   `dv5_base`     | `dv5/RAR`          | $Gli^{dv5}/Gli^{RAR77}$  | -->
<!-- |                  | `dv5/dv5`          | $Gli^{dv5}/Gli^{dv5}$    | -->


# Load necessary libraries and do setup
```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(broom)
library(FSA) # Perform Dunn's test with Bonferroni adjustment
library(latex2exp)
```

# Get Bx-Gal4 data ready

## Load and clean data
```{r, warning=FALSE}
raw_bx_data <- read_excel("data/RNAiscores.xlsx", sheet = 1)

# Reorder and set Genotype factor
raw_bx_data <- raw_bx_data %>%
  mutate(genotype = factor(genotype, levels = c('lacZ', 'sqh-RNAi', 'zip-RNAi')),
         temp = factor(temp))

# Print out summary of the data
str(raw_bx_data)
head(raw_bx_data)
summary(raw_bx_data)
```
# Explore
This section gives some summary statistics and initial visualizations of the dataset.

## Means and Standard deviations of Score
```{r}
summarized_bx_data <- raw_bx_data %>%
  group_by(temp, genotype) %>%
  summarize(mean_score = mean(score, na.rm = TRUE),
            sd_score = sd(score, na.rm = TRUE),
            median_score = median(score, na.rm = TRUE))

# Print out summary of the data
head(summarized_bx_data, n=10)
```


## Plots of Score distributions

<!-- ```{r} -->
<!-- # Boxplot -->
<!-- p <- ggplot(raw_bx_data, aes(x = temp, y = score, fill = genotype)) + -->
<!--   # geom_boxplot(position = position_dodge(width = 1.0), width = 0.8, aes(fill = genotype),  -->
<!--                # color = "black") + -->
<!--   geom_boxplot() -->
<!--   # stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "black", -->
<!--                # position = position_dodge(width = 1.0)) + -->
<!--   # scale_fill_manual(values = c("white", "lightgrey", "darkgrey")) + -->
<!--   labs(title = "Box Plots of Score by Temperature and Genotype", x = "Temp", -->
<!--        y = "Score") + -->
<!--   # geom_jitter(position = position_jitter(width = 0.05), alpha = 0.6) + -->
<!--   theme_minimal() -->
<!-- # Save the plot -->
<!-- # ggsave("figures/hair_scores_boxplots_scaled.tiff", plot = p, width = 10, height = 5, dpi = 300) -->
<!-- p -->
<!-- ``` -->

```{r, fig.height=3, fig.width=7}
# Custom labels for genotype fill
custom_labels <- c(
  "bx-Gal4" = TeX("bx-Gal4"),
  "lacZ" = TeX("UAS-lacZ"),
  "sqh-RNAi" = TeX("UAS-sqh-RNAi"),
  "zip-RNAi" = TeX("UAS-zip-RNAi"),
  "nub-Gal4" = TeX("nub-Gal4")
)

# Set seed for reproducibility
set.seed(2)

# Plot
p <- ggplot(raw_bx_data, aes(x = temp, y = score, fill = genotype)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              alpha = 0.5, shape = 21, color = "black") +  # points with fill
  stat_summary(fun = median, geom = "crossbar", width = 0.7,
               color = "black",
               position = position_dodge(width = 0.8)) +  # median line
  labs(title = "Scores by Temperature and Genotype", x = "Temp", y = "Score", fill = "Genotype") +
  scale_fill_discrete(labels = custom_labels) +
  theme_minimal()
  # theme(legend.position = c(1, 1), legend.justification = c(1.75, 1),
  #       legend.background = element_rect(fill = "white", color = "black", linewidth = 0.1)
  #     ) 
# Optional: Save the plot
# ggsave("figures/bx_temp_and_genotype.tiff", plot = p, width = 7, height = 3, dpi = 300)
# ggsave("figures/bx_temp_and_genotype.svg", plot = p, device="svg", width = 7, height = 3, dpi = 300)

p
```


```{r}
ggplot(raw_bx_data, aes(x = score)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  facet_grid(temp ~ genotype) +
  labs(title = "Histogram of Scores by Temperature and Genotype",
       x = "Score", y = "Count") +
  theme_minimal()
```

The plot above shows the distribution of the scores by temperature in each genotype category. We investigate if these differences are statistically significant in the next section.

# Analysis

ANOVA assumptions are violated (residuals of ANOVA model are not normally distributed). So, we use non-parametric approaches to compare medians.

## Non-parametric Methods: 18 temp

For the 18 temp samples, we use the Kruskal-Wallis test to compare medians followed by a post-hoc test for pairwise differences.

```{r, fig.height=4, fig.width=5}
# Subset data for temp == "18"
data_18temp <- raw_bx_data %>% filter(temp == "18")

# Perform Kruskal-Wallis test
kruskal_result <- kruskal.test(score ~ genotype, data = data_18temp)
kruskal_result
```
```{r}
dunn_result <- dunnTest(score ~ genotype, data = data_18temp, method = "bonferroni")
dunn_result
```

## Non-parametric Methods: 22 temp

For the 22 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_22temp <- raw_bx_data %>% filter(temp == "22")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_22temp, exact=FALSE)
wilcox_result
```
## Non-parametric Methods: 25 temp

For the 25 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_25temp <- raw_bx_data %>% filter(temp == "25")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_25temp, exact=FALSE)
wilcox_result
```
## Non-parametric Methods: 29 temp

For the 29 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_29temp <- raw_bx_data %>% filter(temp == "29")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_29temp, exact=FALSE)
wilcox_result
```

The non-parametric analyses show that all pairings of genotypes within a given temperature have statistically different median scores.


# Get nub-Gal4 data ready

## Load and clean data
```{r, warning=FALSE}
raw_nub_data <- read_excel("data/RNAiscores.xlsx", sheet = 2)

# Reorder and set Genotype factor
raw_nub_data <- raw_nub_data %>%
  mutate(genotype = factor(genotype, levels = c('lacZ', 'sqh-RNAi', 'zip-RNAi')),
         temp = factor(temp))

# Print out summary of the data
str(raw_nub_data)
head(raw_nub_data)
summary(raw_nub_data)
```
# Explore
This section gives some summary statistics and initial visualizations of the dataset.

## Means and Standard deviations of Score
```{r}
summarized_nub_data <- raw_nub_data %>%
  group_by(temp, genotype) %>%
  summarize(mean_score = mean(score, na.rm = TRUE),
            sd_score = sd(score, na.rm = TRUE),
            median_score = median(score, na.rm = TRUE))

# Print out summary of the data
str(summarized_nub_data)
head(summarized_nub_data, n=10)
summary(summarized_nub_data)
```


## Plots of Score distributions

<!-- ```{r} -->
<!-- # Boxplot -->
<!-- p <- ggplot(raw_nub_data, aes(x = temp, y = score, fill = genotype)) + -->
<!--   # geom_boxplot(position = position_dodge(width = 1.0), width = 0.8, aes(fill = genotype),  -->
<!--                # color = "black") + -->
<!--   geom_boxplot() -->
<!--   # stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "black", -->
<!--                # position = position_dodge(width = 1.0)) + -->
<!--   # scale_fill_manual(values = c("white", "lightgrey", "darkgrey")) + -->
<!--   labs(title = "Box Plots of Score by Temperature and Genotype", x = "Temp", -->
<!--        y = "Score") + -->
<!--   # geom_jitter(position = position_jitter(width = 0.05), alpha = 0.6) + -->
<!--   theme_minimal() -->
<!-- # Save the plot -->
<!-- # ggsave("figures/hair_scores_boxplots_scaled.tiff", plot = p, width = 10, height = 5, dpi = 300) -->
<!-- p -->
<!-- ``` -->

```{r, fig.height=3, fig.width=7}
# Plot
p <- ggplot(raw_nub_data, aes(x = temp, y = score, fill = genotype)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
              alpha = 0.5, shape = 21, color = "black") +  # points with fill
  stat_summary(fun = median, geom = "crossbar", width = 0.7,
               color = "black",
               position = position_dodge(width = 0.8)) +  # median line
  labs(title = "Scores by Temperature and Genotype", x = "Temp", y = "Score", fill = "Genotype") +
  scale_fill_discrete(labels = custom_labels) +
  theme_minimal()
  # theme(
  #   legend.position = c(1, 1),
  #   legend.justification = c(4, 1),
  #   legend.background = element_rect(fill = "white", color = "black", linewidth = 0.1)
  # )

# Optional: Save the plot
# ggsave("figures/nub_temp_and_genotype.tiff", plot = p, width = 7, height = 3, dpi = 300)
# ggsave("figures/nub_temp_and_genotype.svg", plot = p, device="svg", width = 7, height = 3, dpi = 300)

p
```


```{r}
ggplot(raw_nub_data, aes(x = score)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  facet_grid(temp ~ genotype) +
  labs(title = "Histogram of Scores by Temperature and Genotype",
       x = "Score", y = "Count") +
  theme_minimal()
```

The plot above shows the distribution of the scores by temperature in each genotype category. We investigate if these differences are statistically significant in the next section.

# Analysis

ANOVA assumptions are violated (residuals of ANOVA model are not normally distributed). So, we use non-parametric approaches to compare medians.

## Non-parametric Methods: 18 temp

For the 18 temp samples, we use the Kruskall-Wallis test to compare medians followed by a post-hoc test for pairwise differences.

```{r, fig.height=4, fig.width=5}
# Subset data for temp == "18"
data_18temp <- raw_nub_data %>% filter(temp == "18")

# Perform Kruskal-Wallis test
kruskal_result <- kruskal.test(score ~ genotype, data = data_18temp)
kruskal_result
```
```{r}
dunn_result <- dunnTest(score ~ genotype, data = data_18temp, method = "bonferroni")
dunn_result
```

## Non-parametric Methods: 22 temp

For the 22 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_22temp <- raw_nub_data %>% filter(temp == "22")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_22temp, exact=FALSE)
wilcox_result
```
## Non-parametric Methods: 25 temp

For the 25 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_25temp <- raw_nub_data %>% filter(temp == "25")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_25temp, exact=FALSE)
wilcox_result
```
## Non-parametric Methods: 29 temp

For the 29 temp samples, we use the Mann-Whitney U test to compare medians.
```{r}
# Subset only the two temp levels
data_29temp <- raw_nub_data %>% filter(temp == "29")

# Mann-Whitney U test (a.k.a. Wilcoxon rank-sum test)
wilcox_result <- wilcox.test(score ~ genotype, data = data_29temp, exact=FALSE)
wilcox_result
```

The non-parametric analyses show that all pairings of genotypes within a given temperature have statistically different median scores. The only exception at the 0.05 level of significance is the comparison of `lacZ - zip-RNAi` for 18 temp which had a p-value of about 0.088.





